Staged Abstract Interpreters

### 2019-02-21 paper/10_intro.tex
add lms support

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. By borrowing the ideas from \note{monadic interpreters}
\cite{DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17,
Sergey:2013:MAI:2491956.2491979} and embedded domain-specific languages
\cite{DBLP:conf/snapl/RompfBLSJAOSKDK15, DBLP:journals/jfp/CaretteKS09,
DBLP:conf/icfp/GibbonsW14}, we develop an approach to fulfill four different
semantics (Figure~\ref{confluence}) in a unified framework. The four semantics
share one same generic interpreter. However, by instantiating the interpreter
differently, it can behave as a concrete interpreter, an abstracter interperter,
or their staged versions, respectively. The key idea to enable this flexibility
is to use abstract type members \note{or type parameters} abstract over concrete/abstract components
(e.g., concrete values or abstract values) of the interpreter, as well as the
binding-time (e.g., static or dyanmic).

Besides the intellectual merit of the confluence diagram, this paper contributes
to a practical problem in static analysis: \textbf{can we speed up a static
analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between precision and efficiency, we argue
that applying meta-programming (particularly, multi-stage programming) with
abstract interpreters is an effective approach to improve the latter one while
fix the former one. Specialization removes the overhead of abstractions
introduced by high-level programming, but not changes the meaning of the program
-- in this case, the abstract semantics implemented by the analyzer. Indeed,
specializing analysis with respect to a program by partial evaluation
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not a new idea, despite the fact that most of the previous
work are ad-hoc on a particular analysis or requiring significant changes to the
analysis. For example, abstract compilation
\cite{Boucher:1996:ACN:647473.727587} requires the whole analyzer to be
rewritten as closure generation from. In this paper, we show a systematic
approach to meta-programming achieves the performance goal with less intrusive
changes; in fact, a case study (Section~\ref{cs_ac}) comparing with abstract
compilation shows, by utilizing type-based stage annotations, the analyzer
program does not need any changes.

@@

\begin{itemize}
  \item We present a unified implementation framework that naturally extends the
    first Futamura projection to abstract interpreters. Starting from a generic
    interface of the interpreter for a tiny functional language
    (Section~\ref{generic_if}), we first show the instantiation of concrete
    interpretation in Section~\ref{unstaged_conc}. Then we lift it to the staged
    version by adding stage annotations to the semantic operations
    (Section~\ref{stagedinterp}); and to the abstract interpreter by applying
    abstraction to the environment, store, and values. Finally, we show the
    combination of the two transformations, dubbed \textit{staged abstract
    interpreters}, can be easily obtained (Section~\ref{sai}).

>>>>

In this paper, we study the confluence of two old ideas --- Futamura projection
and abstract interpretation, but from the perspective of their recent
realizations --- multi-stage programming and definitional abstract interpreters.
To be specific, we present the application of the first Futamura projection on
definitional abstract interpreters, which intersects as staged abstract
interpreters. By borrowing the ideas from \note{monadic interpreters}
\cite{Steele:1994:BIC:174675.178068, DBLP:conf/popl/LiangHJ95, DBLP:journals/pacmpl/DaraisLNH17,
Sergey:2013:MAI:2491956.2491979} and embedded domain-specific languages
\cite{DBLP:conf/snapl/RompfBLSJAOSKDK15, DBLP:journals/jfp/CaretteKS09,
DBLP:conf/icfp/GibbonsW14}, we develop an approach to fulfill four different
semantics (Figure~\ref{confluence}) in a unified framework. The four semantics
share one same generic interpreter. However, by instantiating the interpreter
differently, it can behave as a concrete interpreter, an abstracter interperter,
or their staged versions, respectively. The key idea to enable this flexibility
is to use abstract type members \note{or type parameters} abstract over concrete/abstract components
(e.g., concrete values or abstract values) of the interpreter, as well as the
binding-time (e.g., static or dyanmic).

Besides the intellectual merit of the confluence diagram, this paper contributes
to a practical problem in static analysis: \textbf{can we speed up a static
analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and \textit{efficiency}, we argue
that applying meta-programming (multi-stage programming, in particular) with
abstract interpreters is an effective approach to improve the latter one while
the former one is untouched. Specialization removes the overhead of abstractions
introduced by high-level programming, but not changes the meaning of the program
-- in this case, the abstract semantics implemented by the analyzer. Indeed,
specializing analysis with respect to a program by partial evaluation
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not a new idea, despite the fact that most of the previous
work are ad-hoc on a particular analysis or requiring significant changes to the
analysis. For example, abstract compilation
\cite{Boucher:1996:ACN:647473.727587} requires the whole analyzer to be
rewritten as closure generation from. In this paper, we show a systematic
approach to meta-programming achieves the performance goal with less intrusive
changes; in fact, a case study (Section~\ref{cs_ac}) comparing with abstract
compilation shows, by utilizing type-based stage annotations, the analyzer
program does not need any changes.

@@

\begin{itemize}
  \item We progressively present a unifying construction that naturally extends the
    first Futamura projection to abstract interpreters. Starting from a generic
    interface of the interpreter for a tiny functional language
    (Section~\ref{generic_if}), we first show the instantiation of concrete
    interpretation in Section~\ref{unstaged_conc}. Then we lift it to the staged
    version by adding stage annotations to the semantic operations
    (Section~\ref{stagedinterp}); and to the abstract interpreter by applying
    abstraction to the environment, store, and values. Finally, we show the
    combination of the two transformations, dubbed \textit{staged abstract
    interpreters}, can be easily derived (Section~\ref{sai}).
